Evolutionary Design of Flocking Particles for Collectively Intelligent Problem Solving

### 2014-05-08 docs/paper.tex
final

Many approaches augment the agents with such features as working memory \cite{winder2012role,hu2003particle}. Our paper builds upon the work of Rodriguez & Reggia in \cite{rodriguez2004extending}, who added both working memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal. They were able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.

@@

\section{Discussion}
The evolution designed particles surpassed their human design counterparts through a qualitatively novel mechanism; the human designed team prefered to stay in one or two larger flocks; whereas the high performing evolved teams split into more, smaller flocks and were able to exploit the search space more effectively.

This work described herein represents only the first steps in the application of evolutionary techniques to this problem domain. By mutating and applying selective pressure to the movement component parameters, we were able to show that a machine could optimize the parameters better than a person. This was a form of optimization, but it did not result in novel behaviors. Indeed, since we hard-wired the transitions between the states (allowing for the guarding threshold to evolve) and allowed the evolver to mutate existing movement components in a given state but not to add them (i.e., we pre-defined which components could have non-zero weight for each state), we held constant the "meaning" of each state.

The next step would be to evolve the structure of the FSM itself; not only to allow the evolver to mix-in any component to any state, but to allow it to add new states, delete old states, and change the transitions between the states.

>>>>

Many approaches augment the agents with such features as working memory \cite{winder2012role,hu2003particle}. Our paper builds upon the work of Rodriguez \& Reggia in \cite{rodriguez2004extending}, who added both working memory and a finite state machine (FSM) controller that switched the agent between different sets of movement behaviors according to its local environment and current goal. They were able to demonstrate that a team of such agents was able to solve a resource locate-and-collect problem, and that a team programmed with flocking behaviors outperformed teams of agents that did not influence each other's movements.

@@

\section{Discussion}
A nature-inspired, evolutionary technique was successfully able to design particles that not only exhibited collective, emergent behavior, but also surpassed their human design counterparts through a qualitatively novel mechanism. From epirical observation of the best performing teams, it appears that the human designed team prefered to stay in one or two larger flocks; whereas the high performing evolved teams split into more, smaller flocks and were able to exploit the search space more effectively. Because of the limited time frame in the simulation; this was clearly a better strategy, although additional experiments would be needed to determine the success of these particles over a larger time period.

The computational requirements for performing such an evolution were significant particularly in terms of time. However, parallel computation strategies enabled us to gather initial results. Future work might include MapReduce implementations of the evolutionary computation to distribute tasks across a cluster of computers, as well as longer runs to ensure more significant evolutionary behavior. In order to carry out further experimentation at scale, some larger computational framework would be needed to implement the demands of the simulations.

In particular, the work described in this paper represents only the first steps in the application of evolutionary techniques to the design of particles that exhibit flocking behavior. By mutating and applying selective pressure to the movement component parameters, we were able to show that a machine could optimize bottom-up velocity parameters better than a person, along with some limited state transition logic. However, because the states of the FSM controller were encoded in the simulation, we have not shown the ability to generate novel logical behaviors; simply better collective strategies.

The next steps for this research would be to continue to evolve the structure of the FSM in more meaningful ways. In particular, new states should be generated by the configuration of their individual velocity components as well as other evolved conditional rules; not just the state parameters. Of interest is the opportunity to evolve the transitions between states in a more significant manner, and then to go further and evolve the size and structure of the FSM by adding and removing states.

One essential research question that should be answered is whether or not some form of cross-over is possible in the FSM networks of particle swarms. If so, this will open up more opportunties for searching the optimal FSM space for collective movments. Other research questions might include better fitness functions that utilize the state of the entire world (for example, punishing swarms for allowing enemy agents to collect resources), and speciation for different strategies of evolution.
